### 2019-06-19 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Rien de neuf
Correcting some stuff in Ch4

As explained before in Chapter~\ref{chapter:1}, the future IoT networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use MAB learning algorithms to improve spectral resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks.
We mean by \emph{decentralized} that learning is performed on the device side, and is spread on the totality of the devices in a network.
We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms
firstly extends devices battery autonomy (by reducing the packet-loss-ratio or equivalently, by improving the successful-communication rate),
and secondly does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied settings,
Stochastic MAB learning
% provides an up to $16\%$ gain in terms of successful transmission probabilities, and
has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (two are shown in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}, in left or right).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay, to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

@@

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the \SI{868}{\mega\hertz} ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
\indent
On the opposite, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $D \leq K$ dynamic devices.} than the number of channels}.

@@

\paragraph{$1^{\text{st}}$ - Naive policy: Random Channel Selection} (``\textcolor{darkgreen}{Random}'')
%
We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.
Note that even nowadays this is still the solution implemented for real IoT devices deployed in new LoRaWAN networks.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows
%
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{k=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;k)}_{\text{No one else sent in channel}\; k} \; \underbrace{\Pr(\text{sent in channel}\,k)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

@@

\paragraph{$2^{\text{nd}}$ - (Unachievable) Optimal oracle policy} (``\textcolor{orange}{Optimal}'')
%
We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of the repartition of static devices $(S_k)_k$, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

@@

\paragraph{$3^{\text{rd}}$ - A greedy approach of the oracle strategy} (``\textcolor{deeppurple}{Good sub-optimal}'')
%
We propose a \emph{sequential} approximation of the optimal policy:
the third solution is a sub-optimal naive policy, simple to set up, but also unpractical as it also needs an oracle.
End-devices are iteratively inserted in the channels with the lowest load (\ie, the index $k$ minimizing $S_k + D_k(\tau)$ at global time step $\tau$). Once the number of devices in each channel is computed, the probability of sending successfully a message is also given by equation \eqref{eq:41:prob_col}.
This is the policy that would be used by dynamic devices if they were inserted one after the other, and if they had a perfect knowledge of the channel loads.

>>>>

As explained before in Chapter~\ref{chapter:1}, the future IoT networks will require to support more and more communicating devices.
In this section, we show that intelligent devices in unlicensed bands can use MAB learning algorithms to improve spectral resource exploitation.
%
We evaluate the performance of two classical MAB learning algorithms, \UCB{} and Thomson Sampling, to handle the decentralized decision-making of Spectrum Access, applied to IoT networks.
We mean by \emph{decentralized} that learning is performed on the device side, and is spread on the totality of the devices in a network.
We also evaluate the learning performance when the number of intelligent end-devices grows.
%
We show that using learning algorithms
firstly extends devices battery autonomy (by reducing the packet-loss-ratio or equivalently, by improving the successful-communication rate),
and secondly does help to fit more devices in such networks, even when all end-devices are intelligent and are dynamically switching from channel to channel.
In the studied settings, stochastic MAB learning
% provides an up to $16\%$ gain in terms of successful transmission probabilities, and
has near optimal performance even in non-stationary settings with a majority of intelligent devices.

@@

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic, more and more fluctuating and unpredictable.

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (two are shown in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}, in left or right).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay, to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of orthogonal wireless channels, and a unique PHY layer configuration (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

We note that MAB learning has also started to be applied to also optimize the PHY layer parameters, see for instance \cite{KerkoucheAlami18} following our article \cite{Bonnefoi17},
but in this thesis we chose to restrict to the spectrum access problem and thus we only MAB models where the arms correspond to orthogonal frequency bands.

@@

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the \SI{868}{\mega\hertz} ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
\indent
On the opposite, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $M = D \leq K$ dynamic devices, referred to as \emph{players}.} than the number of channels}.

@@

\paragraph{$1^{\text{st}}$ - Naive policy: Random Channel Selection} (``\textcolor{darkgreen}{Random}'')

We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.
Note that even nowadays this is still the solution implemented for real IoT devices deployed in new LoRaWAN networks.

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$,
so probability of successful transmission is computed as follows

\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{k=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;k)}_{\text{No one else sent in channel}\; k} \; \underbrace{\Pr(\text{sent in channel}\,k)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

@@

\paragraph{$2^{\text{nd}}$ - (Unachievable) Optimal oracle policy} (``\textcolor{orange}{Optimal}'')

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of the repartition of static devices $(S_k)_k$, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

@@

\paragraph{$3^{\text{rd}}$ - A greedy approach of the oracle strategy} (``\textcolor{deeppurple}{Good sub-optimal}'')

We propose a \emph{sequential} approximation of the optimal policy:
the third solution is a sub-optimal naive policy, simple to set up, but also unpractical as it also needs an oracle.
End-devices are iteratively inserted in the channels with the lowest load (\ie, the index $k$ minimizing $S_k + D_k(\tau)$ at global time step $\tau$). Once the number of devices in each channel is computed, the probability of sending successfully a message is also given by equation \eqref{eq:41:prob_col}.
This is the policy that would be used by dynamic devices if they were inserted one after the other, and if they had a perfect knowledge of the channel loads.

### 2019-06-19 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main.tex
Correcting some stuff in Ch4

In this section, we present a demonstration showcased at the International Conference on Telecommunication (ICT) in June $2018$ \cite{Besson2018ICT,Besson2019WCNC}, implementing a proof-of-concept (PoC) of the model introduced in Section~\ref{sec:4:firstModel}.
%
As far as we know, this is the first demonstration of running learning algorithms on the IoT device side, in real radio conditions, as highlighted it in our paper \cite{MoyBesson2019}.

>>>>

In this section, we present a demonstration showcased at the International Conference on Telecommunication (ICT) in June $2018$ \cite{Besson2018ICT,Besson2019WCNC}, implementing a proof-of-concept (PoC) of the model introduced in Section~\ref{sec:4:firstModel}.
%
As far as we know, this is the first demonstration of running learning algorithms on the end-device side for Internet of Things networks, in real radio conditions, as highlighted it in our paper \cite{MoyBesson2019}.

### 2019-06-19 2-Chapters/4-Chapter/chapter4.tex
Correcting some stuff in Ch4

We already explained in Chapter~\ref{chapter:1} that
unlicensed bands are more and more used and considered for mobile and LAN communication standards (WiFi, LTE-U), and for Internet of Things (IoT) standards for short-range (ZigBee, Z-Wave, Bluetooth) and long-range (LoRaWAN, SIGFOX, Ingenu, Weightless) communications \cite{Centenaro16}.
This heavy use of unlicensed bands, in particular with the expected exponential growth of the number of IoT devices, will cause performance drop, due to radio collisions that could even compromise IoT promises.

Efficient Medium Access (MAC) policies allow devices to avoid interfering traffic and can significantly reduce the spectrum contention problem in unlicensed bands.
As end-devices battery life is a key constraint of IoT networks,
and as IoT networks are decentralized (\ie, the devices initiate transmissions),
this leads to IoT protocols using as low signaling overhead as possible and simple ALOHA-based mechanisms.
%
In this chapter, we analyze the performance of Multi-Armed Bandits (MAB) algorithms, that could be used in combination with a time-frequency slotted ALOHA-based protocol.
We highlight that even without changing anything on the level of IoT standards, our proposal is just an add-on capability that can be used on a unit-per-unit basis.
We consider the Upper-Confidence Bound (\UCB) \cite{Auer02}, and the Thompson-Sampling (TS) algorithms \cite{Thompson33,AgrawalGoyal11,
Kaufmann12Thompson}, for the first model. For the demonstration as well as for the second model, without loss of generality, we preferred to focus on heuristics based on the simplest algorithm (\ie, \UCB), to give a clear presentation of the different ideas explored to solve the problem of learning in order to retransmit efficiently.

As detailed in Chapter~\ref{chapter:1},
MAB learning has already been proposed in Cognitive Radio (CR) \cite{Mitola99,Haykin05}, and in particular, for sensing-based Dynamic Spectrum Access (DSA) in licensed bands \cite{Jouini10}.
For example,
proof-of-concepts like \cite{kumar2016two} have proven the capability of such approaches on real radio signals for OSA,
and \cite{Maghsudi16} shows how MAB learning can be applied for small cell management in licensed 5G networks.
Some analysis on real radio measurements made for HF ionospheric channels have also proven that solutions based on MAB learning is appropriate and solves efficiently this kind of decision-making problems on real-world wireless signals \cite{Melian15}.
Recent works show that stationary MAB algorithms work well to solve reinforcement learning models that represent accurately real-world radio problems.
Recently, TS and \UCB{} algorithms have been used for improving the spectrum access in (unlicensed) WiFi networks, for instance by \cite{Toldov16} or \cite{Wilhelmi19collaborative,Wilhelmi19potential}.
% However, even with only one dynamic user using the learning algorithm, the background traffic or the traffic of the other devices is never really stationary or \iid{}.

@@

% ----------------------------------------------------------------------------
\section{Test-bed implementation of our model for real-world validation}
\label{sec:4:gnuradio}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------------
\section{Extending the model to account for retransmissions}
\label{sec:4:retransmissions}
% ----------------------------------------------------------------------

>>>>

After the first chapters that presented the model of MAB, we go back to the initial problems studied in this thesis, and thus we focus on Internet of Things (IoT) networks.
%
As explained in the introduction in Chapter~\ref{chapter:1},
unlicensed bands are more and more used and considered for mobile and LAN communication standards (WiFi, LTE-U), and for Internet of Things (IoT) standards for short-range (ZigBee, Z-Wave, Bluetooth) and long-range (LoRaWAN, SIGFOX, Ingenu, Weightless) communications \cite{Centenaro16}.
This heavy use of unlicensed bands, in particular with the expected exponential growth of the number of IoT devices, will cause performance drop, due to radio collisions that could even compromise IoT promises.

Efficient Medium Access (MAC) policies allow devices to avoid interfering traffic and can significantly reduce the spectrum contention problem in unlicensed bands.
As end-devices battery life is a key constraint of IoT networks,
and as IoT networks are decentralized, because the devices initiate transmissions,
this leads to IoT protocols using as low signaling overhead as possible and simple ALOHA-based mechanisms.
%
In this chapter, we analyze the performance of Multi-Armed Bandits (MAB) algorithms, that could be used in combination with a time-frequency slotted ALOHA-based protocol.
We highlight that even without changing anything on the level of IoT standards, our proposal is just an add-on capability that can be used on a unit-per-unit basis.
We consider the Upper-Confidence Bound (\UCB) \cite{Auer02}, and the Thompson-Sampling (TS) algorithms \cite{Thompson33,AgrawalGoyal11,
Kaufmann12Thompson}, for the first model. For the demonstration as well as for the second model, without loss of generality, we preferred to focus on heuristics based on the simplest algorithm (\ie, \UCB), to give a clear presentation of the different ideas explored to solve the problem of learning in order to retransmit efficiently.

% As detailed in Chapter~\ref{chapter:1},
% MAB learning has already been proposed in Cognitive Radio (CR) \cite{Mitola99,Haykin05}, and in particular, for sensing-based Dynamic Spectrum Access (DSA) in licensed bands \cite{Jouini10}.
% For example,
% proof-of-concepts like \cite{kumar2016two} have proven the capability of such approaches on real radio signals for OSA,
% and \cite{Maghsudi16} shows how MAB learning can be applied for small cell management in licensed 5G networks.
% Some analysis on real radio measurements made for HF ionospheric channels have also proven that solutions based on MAB learning is appropriate and solves efficiently this kind of decision-making problems on real-world wireless signals \cite{Melian15}.
% Recent works show that stationary MAB algorithms work well to solve reinforcement learning models that represent accurately real-world radio problems.
% Recently, TS and \UCB{} algorithms have been used for improving the spectrum access in (unlicensed) WiFi networks, for instance by \cite{Toldov16} or \cite{Wilhelmi19collaborative,Wilhelmi19potential}.
% % However, even with only one dynamic user using the learning algorithm, the background traffic or the traffic of the other devices is never really stationary or \iid{}.

@@

\newpage  % WARNING ?
% ----------------------------------------------------------------------------
\section{Test-bed implementation of our model for real-world validation}
\label{sec:4:gnuradio}
% ----------------------------------------------------------------------

@@

\newpage  % WARNING ?
% ----------------------------------------------------------------------------
\section{Extending the model to account for retransmissions}
\label{sec:4:retransmissions}
% ----------------------------------------------------------------------
